Linear Stochastic Approximation- Constant Step-Size and Iterate Averaging

### 2017-05-08 related.tex
updated version

\section{Related Work}
\subsection{Linear Regression}
Let $(x_t,y_t)\in \R^n\times \R,\,t\geq 1$ be $i.i.d$ such that $\EE{\norm{x_t}^2}$ and $\EE{y_t^2}$ are finite. Further, we assume bounded data i.e., $\norm{(x_t,y_t)}^2\leq B$ for some positive real number $B>0$. Here, $x_t,\,t\geq 1$ are the features and $y_t$ are real labels and the linear regression problem constitutes minimizing the quadratic loss function $f(\theta)=\EE{\left(\ip{x_t,\theta}-y_t\right)^2}$. The stochastic gradient descent scheme with constant step size $\alpha>0$ to minimize $f(\theta)$ can be given as below:
\begin{align}\label{linreg}
\theta_t=\theta_{t-1}-\alpha x_t\left(\ip{x_t,\theta_{t-1} -y_t}\right)
\end{align}
Here $\nabla f(\theta_{t-1})=x_t\left(\ip{x_t,\theta_{t-1} -y_t}\right)$ is the \emph{stochastic gradient} of the objective $f(\theta)$. It follows from the $i.i.d$ assumption on $(x_t,y_t)$ that $A_t=x_t x_t^\top$ and $b_t=y_t x_t$ are also $i.i.d$, and with this identification one can see that \eqref{linreg} is of the form in \eqref{conststep}. In this case $A_P$ is SPD.
In the linear regression setting, the error of a parameter $\theta$ is measured as the difference between the loss at the said parameter and the loss at the optimum, which has the following form
\begin{align*}
f(\theta)-f(\ts)=(\theta-\ts)^\top A_P (\theta-\ts)=\EE{\norm{\theta-\ts}^2_{A_P}}
\end{align*}

\subsection{Reinforcement Learning}
\subsection{GTD}
\todoch{Use LiHong Li's GTD results, choice of the basis $U$ etc}

>>>>

\section{Applications in Reinforcement Learning}
\paragraph{Definitions:} We are presented with data in the form of an $i.i.d$ sequence $(\phi_t,\phi'_t,r_t)\in \R^d\times\R^d\times \R$. Let $\gamma \in (0,1)$ be a fixed constant, we define $\Delta_t\eqdef \phi_t\phi_t^\top-\gamma \phi_t\phi_t^{'\top}$, $C_t=\phi_t\phi_t^\top$ and $b_t=\phi_r r_t$. In what follows, $\mu_t>0$ is an \emph{importance} sampling factor whose aim is to correct for mismatch in the (behavior) distribution with which the data was collected and the (target) distribution with respect to which one wants to learn. A factor $\mu_t=1,\,\forall t\geq 0$ will mean that no correction is required\footnote{This is known as the \emph{on-policy} case where the behavior is identical to the target. The general setting where $\mu_t>0$ is known as \emph{off-policy}.}.
\begin{table}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|c|c|c|c|c|}\hline
Algorithm& Update &Stepsize &Rate \\ \hline
TD(0)& $\begin{aligned}\theta_t&=\theta_{t-1}+\alpha_t (b_t -\Delta_t\theta_{t-1})\\ \thh_t&=\frac{1}{t+1}\sum_{s=0}^t \theta_t\end{aligned}$ &$\begin{aligned}\alpha_t&=O(\frac1{t})^\delta,\\ &\delta\in(1/2,1)\end{aligned}$ &$\EE{\norm{\eh_t}}\leq O(\frac{1}{\sqrt{t}})$ for $\delta\ra 1$ \\ \hline
GTD& $\begin{aligned} y_{t}&=y_{t-1}+\beta_t(\mu_t b_t -\mu_t \Delta_t \theta_{t-1}- Q_t y_{t-1})\\ \theta_t &=\theta_{t-1}+\alpha_t (\mu_t A^\top_t y_{t-1})\end{aligned}$ & $\begin{aligned}&\beta_t=\eta \alpha_t,\sum_{t\geq 0}\alpha_t=\infty, \\ &\sum_{t\geq 0}\alpha^2_t<\infty \end{aligned}$ & $e_t\ra 0$ w.p.1 \\\hline
TDC& $\begin{aligned} y_{t}&=y_{t-1}+\beta_t(\mu_t b_t -\mu_t \Delta_t \theta_{t-1}-C_t y_{t-1})\\ \theta_t &=\theta_{t-1}+\alpha_t (\mu_t b_t-\mu_t\Delta_t\theta_{t-1}+ (\Delta^\top_t-C_t)y_{t-1})\end{aligned}$ & $\begin{aligned}&\beta_t,\alpha_t,\frac{\alpha_t}{\beta_t} \ra 0,~\text{as}~t\ra \infty,\\& \sum_{t\geq 0}\alpha_t=\infty, \sum_{t\geq 0}\beta_t=\infty, \\ &\sum_{t\geq 0}\alpha^2_t<\infty, \sum_{t\geq 0}\beta^2_t<\infty \end{aligned}$ & $e_t\ra 0$ w.p.1 \\\hline
GTD& $\begin{aligned} y_{t}&=\Pi_Y \left(y_{t-1}+\alpha_t(\mu_t b_t -\mu_t \Delta_t \theta_{t-1}-Q_t y_{t-1})\right)\\ \theta_t &=\Pi_{\Theta}\left(\theta_{t-1}+\alpha_t (\mu_t A_t^\top y_{t-1})\right)\\ \thh_t&=\frac{\sum_{s=0}^t \alpha_s\theta_s}{\sum_{s=0}^t \alpha_s}\end{aligned}$ & $\alpha_t =O(\frac{1}{\sqrt{t}})$ & $\norm{A_P\eh_t-b}^2 \leq O(\frac{1}{\sqrt{t}})$ w.h.p\\\hline

\end{tabular}
}
\label{tb:rates}
\end{table}
\paragraph{Related Work in RL} The various temporal difference learning algorithms can be written as in \Cref{table:td}. The TD(0) algorithm is an on-policy algorithm ($\mu_t=1$) and is unstable in the general {off-policy} setting. The gradient temporal difference (GTD) algorithm \cite{} is stable in off-policy setting as well. Recent work by \cite{korda-prashanth} derive finite time bounds for the TD(0) for two different cases of step-sizes one without PR averaging and other with PR averaging. In the case without PR averaging, \cite{korda-prashanth} consider step-sizes of the form $\frac{c_0c}{c+t}$, where $c_0$ and $c$ are problem dependent positive constants, to dervie a rate of $O(\frac{1}{t})$ for the MSE.  In the case with PR averaging (row $1$ of \Cref{tb:rates}), \cite{korda-prashanth} consider step-size of the from $c_0\left(\frac{c}{c+t}\right)^\delta,\,\delta \in (1/2,1)$ where $c_0, c$ are problem independent positive constants, and show that this choice yeilds a rate of $O(\frac{1}{t})$ for $\delta \ra 1$ and $t\geq t_0$ (where $t_0$ is a problem dependent time instant). A possible limiting feature of the results by \cite{korda-prashanth} is that either the problem knowledge is required or the bounds hold only after a problem dependent time. A centered TD(0) variant is also considered by \cite{korda-prashanth}, however, a projection operator is required to keep the iterates bounded. Also, the bounds by \cite{korda-prashanth} don't have an explicit bias variance decomposition. The works on GTD and TDC by \cite{gtdfirst,tdc} show only asymptotic convergence w.p.1. Recently, the work by \cite{dtdmp} shows a rate of $\frac{1}{\sqrt{t}}$ for GTD with use of projection operation, and also proposed a variant namely GTD-Mirror-prox based on stochastic mirror-prox ideas which achieves $\frac{1}{\sqrt{t}}$ rate without use of projection. In comparison, we don't use of projections, our bounds hold for all time $t$, and the bounds are dependent on the problem instance only to the extent of choosing a step-size that leads to stable behavior, in which case we show that $O\frac{1}{t}$ rate holds (the scaling factor hidden in the $O$ expression is problem dependent). Further, our results provide explicit bias variance decomposition.

### 2017-05-08 stepsizes.tex
updated version

\section{Choice of Step-sizes}
We now argue, in a heuristic manner, as to why constant step-size with PR-averaging might be a principled idea, especially in TD type algorithms. Consider the following LSA with time varying step-sizes $\alpha_t>0$
\begin{align}
\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1}).
\end{align}
It is known from asymptotic results by \cite{borkarbook,kushner}\todoch{Check about Kushner} that in order to handle the noise the step-sizes have to be diminishing i.e., $\alpha_t\ra$, unsummable i.e., $\sum_{t \geq 0}\alpha_t=\infty$ and square summable $\sum_{t\geq 0}\alpha_t^2<\infty$. We now look at two specific diminishing step-size sequences, to illustrate the issues that the diminishing step-sizes presents us with. In what follows, we assume that the maximum eigevalue of $\EE{A_t}$ is less that $1$.
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item \textbf{$\alpha_t=\frac{1}{t}$\footnote{While $\frac{c}{t}$, $c\geq 0$ is more general, it is enough to consider $\frac1t$ by accommodating the multiple $c$ as part of the matrix $A_t$}} is known yield a rate of $\frac{1}{t^{\lambda_{\min}}}$, where $\lambda_{\min}$ is the smallest real part of the eigen values of the matrix $\EE{A_t}$.
\item \textbf{$\alpha_t=\frac{c}{c+t}$} is another example, where the step size decays at a rate $\frac{1}{t/c}$. The rate of decay of error in this case turns out to be $\frac{c^2\sigma_b^2}{t}$, whenever $c>\frac{1}{\lambda_{\min}}$, which can be ensured only when $\lambda_{\min}$ is known.
\end{itemize}
In contrast, in the case of constant step-size with PR-averaging, not knowing the problem instance can result in instability, and when we manage to choose a step-size that results in stable behavior, the rates obtained in \Cref{th:pdrate,cor:as} hold for it. Using this observation, the constant-step size can be set in a more prinicipled manner as showing the pseudocode below.
\begin{algorithm}[H]
\caption*{Automatic Tuning of Constant Step-Size}
\begin{algorithmic}[1]
\STATE{Initialize: $\theta_0$, $\alpha=1$, $i=1$, $a_0=0$}
\FOR{$t=1,2,\ldots, $}
\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1})$}
\IF{$mod(t,k)==0$}
\STATE{$a_{i}=\norm{\theta_t}$}
\IF{$Is-stable(a_0,\ldots,a_i)==0$}
\STATE{$\alpha=\alpha/2$}
\ENDIF
\STATE{$i=i+1$}
\ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{comment}
In order to calculate the rates we make use of the following asymptotic result by \cite{borkarbook}.
\begin{theorem}\label{th:borkar}
Let $\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1})$, where $\theta_t,b_t\in\R^d$ and $A_t\in\R^{\dcd}$ that satisfy the conditions in \Cref{assmp:lsa} and $\sum_{t\geq 0}\alpha_t=\infty$, $\sum_{t\geq 0}\alpha^2_t<\infty$. Let $t_{ode}(k)=\eqdef \sum_{i=0}^{k-1}\alpha_n$ and let $\tilde{\theta}(t)$ be the trajectory of the ordinary differential equation $\dot{\theta}(t)=\EE{b_t}-A_P\theta(t)$. Then it follows that
\begin{align*}
\lim_{k\ra\infty}\sup\norm{\tilde{\theta_t(t_{ode}(k))}-\theta_t} \ra 0
\end{align*}
\end{theorem}

>>>>

\section{Automatic Tuning of Step-Sizes}
\begin{comment}
We now argue, in a heuristic manner, as to why constant step-size with PR-averaging might be a principled idea, especially in TD type algorithms. Consider the following LSA with time varying step-sizes $\alpha_t>0$
\begin{align}
\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1}).
\end{align}
It is known from asymptotic results by \cite{borkarbook,kushner}\todoch{Check about Kushner} that in order to handle the noise the step-sizes have to be diminishing i.e., $\alpha_t\ra$, unsummable i.e., $\sum_{t \geq 0}\alpha_t=\infty$ and square summable $\sum_{t\geq 0}\alpha_t^2<\infty$. We now look at two specific diminishing step-size sequences, to illustrate the issues that the diminishing step-sizes presents us with. In what follows, we assume that the maximum eigevalue of $\EE{A_t}$ is less that $1$.
\begin{itemize}[leftmargin=*, before = \leavevmode\vspace{-\baselineskip}]
\item \textbf{$\alpha_t=\frac{1}{t}$\footnote{While $\frac{c}{t}$, $c\geq 0$ is more general, it is enough to consider $\frac1t$ by accommodating the multiple $c$ as part of the matrix $A_t$}} is known yield a rate of $\frac{1}{t^{\lambda_{\min}}}$, where $\lambda_{\min}$ is the smallest real part of the eigen values of the matrix $\EE{A_t}$.
\item \textbf{$\alpha_t=\frac{c}{c+t}$} is another example, where the step size decays at a rate $\frac{1}{t/c}$. The rate of decay of error in this case turns out to be $\frac{c^2\sigma_b^2}{t}$, whenever $c>\frac{1}{\lambda_{\min}}$, which can be ensured only when $\lambda_{\min}$ is known.
\end{itemize}
In contrast, in the case of constant step-size with PR-averaging, not knowing the problem instance can result in instability, and when we manage to choose a step-size that results in stable behavior, the rates obtained in \Cref{th:pdrate,cor:as} hold for it. Using this observation, the constant-step size can be set in a more prinicipled manner as showing the pseudocode below.
\end{comment}
In \Cref{sec:uniform} we showed a negative result that for the general class of LSAs, it is not possible to pick a fixed uniform step-size that can be used for any instance of the class. Further,
recall from our discussion in \Cref{sec:mainresults} that as $\alpha\ra\alpha_P$ (or $\alpha\ra\alpha_{P_U}$), the MSE blows up and for $\alpha>\alpha_P$ the LSA is unstable because the spectral norm of the random matrix $I-\alpha A_t$ is no longer less that unity. Combining these facts, we conclude that if the step-size is incorrect for a particular instance then we would observe instability, a fact that can be used to reset the step-size to a smaller value, and this procedure can be continued in an iterative fashion to find the right choice of step-size for a given problem. We formalize this idea in the form of a step-size tuning algorithm in \Cref{alg:tuning}

\begin{algorithm}[H]
\caption*{Automatic Tuning of Constant Step-Size}
\begin{algorithmic}[1]
\STATE{Initialize: $\theta_0$, $\alpha=1$, $i=1$, $\rho_0=0$}
\FOR{$t=1,2,\ldots, $}
\STATE{$\theta_t=\theta_{t-1}+\alpha (b_t-A_t\theta_{t-1})$}
\IF{$t (mod k)==0$}
\STATE{$\rho_{i}=\norm{\theta_t}$}
\IF{IsUnstable($\{\rho_j,j=1,\ldots,i\}$)==1}
\STATE{$\alpha=\alpha/2$}
\ENDIF
\STATE{$i=i+1$}
\ENDIF
\ENDFOR
\end{algorithmic}
\label{alg:tuning}
\end{algorithm}
In \Cref{alg:tuning} IsUnstable$(\rho_j,j=1,\ldots,i\}$) is a routine whose input is $\norm{\theta_t}$ measured at time instances $k$ apart. It is easy to show from Gronwall type arguments that the $\norm{\theta_t}\leq (1-exp{\beta t})$, where $\beta>0$ implies stability and $\beta<0$ imples instability. From this Gronwall based inequality, we observe that when the iterates are stable the ratio $\frac{\rho_{k+1}}{\rho_k},k=1,\ldots,i-1$ should be `roughly' decreasing (because there is the iterates are persistently corrupted by noise) and should converge to $1$. This condition on the sequence $\frac{\rho_{k+1}}{\rho_k}$ can be checked in practice with appropriate tolerances.
\begin{comment}
In order to calculate the rates we make use of the following asymptotic result by \cite{borkarbook}.
\begin{theorem}\label{th:borkar}
Let $\theta_t=\theta_{t-1}+\alpha_t(b_t-A_t\theta_{t-1})$, where $\theta_t,b_t\in\R^d$ and $A_t\in\R^{\dcd}$ that satisfy the conditions in \Cref{assmp:lsa} and $\sum_{t\geq 0}\alpha_t=\infty$, $\sum_{t\geq 0}\alpha^2_t<\infty$. Let $t_{ode}(k)=\eqdef \sum_{i=0}^{k-1}\alpha_n$ and let $\tilde{\theta}(t)$ be the trajectory of the ordinary differential equation $\dot{\theta}(t)=\EE{b_t}-A_P\theta(t)$. Then it follows that
\begin{align*}
\lim_{k\ra\infty}\sup\norm{\tilde{\theta_t(t_{ode}(k))}-\theta_t} \ra 0
\end{align*}
\end{theorem}
